Policy optimization for dynamic spatiotemporal systems \hspace{1em} \insertframenumber

### 2014-11-02 pot/LaTeX/skewnormal.tex
update results file with new plots

The paper is organized as follows.
In section \ref{s:model}, we describe the skew-$t$ process with partitioning and thresholding.
The computing is described in section \ref{s:comp}.
In section \ref{s:simstudy}, we present a simulation study that examines the predictive capabilities of this model compared with a na{\"i}ve Gaussian method.
We then compare our method to Gaussian and max-stable methods with a data analysis of ozone measurements from the eastern US in section \ref{s:analysis}.
The final section provides brief discussion and direction for future research.

\section{Statistical model}\label{s:model}
Let $Y_t(\bs)\in \calR$ be the observed value at location $\bs$ and timepoint $t$.
To avoid bias in estimating tail parameters, we model censored data
\beq\label{Yt}
  \Yt_t(\bs) = \left\{
          \begin{array}{ll}
            Y_t(\bs) & Y_t(\bs)>T \\
            T & Y_t(\bs)\le T
          \end{array}
        \right.
\eeq
where $T$ is a pre-specified threshold.
Then, assuming the full data are observations from a skew-$t$ process, we update values censored below the threshold using standard Bayesian missing data methods as described in Section \ref{s:comp}.

\subsection{Skew-$t$ process}\label{s:data}
Many types of data demonstrate some level of skewness, and therefore should be modeled with distributions that allow for non-symmetry.
The skew-elliptical family of distributions provides models that are mathematically tractable while introducing additional paremeters to account for non-symmetric data \citep{Genton2004}.
One member of this family, is the skew-$t$ distribution \citep{Azzalini2003}.
\citet{Zhang2010} show that the skew-$t$ can be expressed as a hierarchical model
\begin{align} \label{eq:fullmodel}
  Y_t(\bs) &= X_t(\bs) \beta + \alpha |z_t| + \sigma_t v_t(\bs)
\end{align}
where $\alpha \in \calR$ controls the skewness, $z_{t} \ind N(0, \sigma^2_t)$ are a random effect from a half-normal distribution (see appendix A.4), $v_t(\bs)$ is a Gaussian process with mean zero, variance one, \Matern correlation, and $\sigma^2_t \iid \text{IG}(a, b)$.
When marginalizing over the $z_t$ and $\sigma^2_t$ terms,
\begin{align*}
  Y_t(\bs) \sim \text{skew-}t (\mu, \Sigma^*, \alpha, \text{df} = 2a)
\end{align*}
where $\mu$ is the location, $\Sigma^* = \frac{ b }{ a } \Sigma$, $\Sigma$ is a \Matern covariance matrix, and $\alpha \in \calR$ controls the skewness.
The skew-$t$ process is desirable because of its flexible tail that is controlled by the skewness parameter, $\alpha$, and the degrees of freedom, $2a$.
Furthermore, if $\bY$ follows a multivariate skew-$t$ distribution, the marginal distributions also follow a skew-$t$ distribution \citep{Azzalini2003}.

\subsection{Extremal dependence}
One common measure of extremal spatial dependence is the extremal coefficient which describes the pairwise dependence between spatial locations \citep{Smith1990}.
Consider a spatial process $Y(\bs) \in \calR^n$ observed at locations $s \in \calD \subset \calR^2$.
Then the bivariate extremal coefficient, $\theta(\bs_i, \bs_j) \in [1, 2]$, is defined as
\begin{align}
  \Pr(Y(\bs_i) < c, Y(\bs_j) < c) = \Pr(Y(\bs_i) < c)^{\theta(\bs_i, \bs_j) }. \label{eq:extremalcoef}
\end{align}
One way to characterize the dependence over the entire set of spatial locations is to calculate all of the pairwise extremal coefficients.
Although this method provides information regarding the spatial structure of the observations, it does not fully characterize the joint spatial dependence.

Another measure of extremal dependence is the $\chi$ statistic.
The $\chi$ statistic for the upper tail is given by
\begin{align*}
  \chi = \lim_{c \rightarrow \infty} \Pr(Y(\bs_1) > c | Y(\bs_2) > c),
\end{align*}
In a stationary spatial process, we can write the $\chi$ coefficient as
\begin{align*}
  \chi(\bh) = \lim_{c \rightarrow \infty} \Pr(Y(0) > c | Y(\bh) > c)
\end{align*}
where $\bh = ||\bs_1 - \bs_2||$.
If $\chi(\bh) = 0$, then observations are asymptotically independent at distance $\bh$.
For Gaussian processes, $\chi(\bh) = 0$ regardless of the the distance, so they are not suitable for modeling spatially dependent extremes.
However, for the process described in Section \ref{s:data}, $\chi(\bh) > 0$ \citep{Padoan2011}.

\subsection{Random daily partition}\label{s:part}
One problem with the spatial skew-$t$ process is that all sites are asymptotically dependent regardless of their spatial separation.
This occurs because all observations, both near and far, share the same $z_t$ and $\sigma^2_t$ terms.
We handle this problem with a daily random partition similar to \citet{Kim2005} that allows $z_t$ and $\sigma^2_t$ to vary by site.
The model then becomes
\begin{align}
  Y_t(\bs) &= X_t(\bs) \beta + \alpha |z_t(\bs)| + \sigma_t(\bs) v_t(\bs).
\end{align}
In this extension of (\ref{eq:fullmodel}), $z_t(\bs)$ and $\sigma_t(\bs)$ are allowed to vary by site.
Their spatial variation is determined by the random partition model defined below.
Consider a set of daily spatial knots $\bw_{tk} \sim \text{Uniform}(\calD)$ that define a random daily partition $P_{t1}, \ldots, P_{tK}$ of the spatial domain of interest $\calD \subset \calR^2$ such that
\begin{align*}
  P_{tk} = \{ \bs : k = \argmin_\ell || \bs - \bw_{t\ell} || \}.
\end{align*}
So, for $\bs \in P_{tk}$, let
\begin{align}
  z_t(\bs) &= z_{tk} \label{eq:sitez}\\
  \sigma^2_t(\bs) &= \sigma^2_{tk} \label{eq:sitesig}.
\end{align}
Then within each partition, $Y_t(\bs)$ follows the distribution given in (\ref{eq:fullmodel}).
When incorporating the random daily partition, the $\chi$ statistic becomes
\begin{align}
  \chi(\bh) = \lim_{c \rightarrow \infty} \pi(\bh) \chi(\bh)
\end{align}
where $\pi(h)$ is the probability that two sites separated by distance $h$ are in the same partition.
Figure \ref{fig:simchi} shows $\chi(h)$ for different sample quantiles from simulated datasets where $\bs \in [0, 1] \times [0, 1]$.
These plots show how the partitioning allows for short range extremal dependence while reducing extremal dependence as the distance between sites increases.
\begin{figure}
  \includegraphics[width=\linewidth]{plots/sim-chi.pdf}
  \caption{$\chi(h)$ for simulated datasets.}
  \label{fig:simchi}
\end{figure}

Asymptotic dependence is eliminated as $h$ increases, because
\begin{align}
  \lim_{h \rightarrow \infty} \chi(\bh) = \lim_{h \rightarrow \infty} \pi(\bh) \chi(\bh) = 0.
\end{align}
A proof of this can be found in Appendix A.3.

>>>>

The paper is organized as follows.
In section \ref{s:spatial}, we describe a spatial skew-$t$ process with partitioning and thresholding.
In section \ref{s:temporal}, we provide a way to model temporal correlation.
The computing is described in section \ref{s:comp}.
In section \ref{s:simstudy}, we present a simulation study that examines the predictive capabilities of this model compared with a na{\"i}ve Gaussian method.
We then compare our method to Gaussian and max-stable methods with a data analysis of ozone measurements from the eastern US in section \ref{s:analysis}.
The final section provides brief discussion and direction for future research.

\section{Spatial model}\label{s:spatial}
To avoid bias in estimating tail parameters, we model censored data.
Let
\beq\label{Yt}
  \tilde{Y}(\bs) = \delta(\bs) Y(\bs) + (1 - \delta(\bs)) T
\eeq
be the censored observation at site $\bs$ where $Y(\bs)$ is the uncensored observation, $\delta(\bs) = I[Y(\bs) > T]$, and $T$ is a pre-specified threshold value.
Then, assuming the uncensored data are observations from a skew-$t$ process, we update values censored below the threshold using standard Bayesian missing data methods as described in Section \ref{s:comp}.

\subsection{Skew-$t$ process}\label{s:data}
Many types of data demonstrate some level of skewness, and therefore should be modeled with distributions that allow for non-symmetry.
The skew-elliptical family of distributions provides models that are mathematically tractable while introducing additional paremeters to account for non-symmetric data \citep{Genton2004}.
One member of this family, is the skew-$t$ distribution \citep[see Appendix A.5]{Azzalini2003}.
\citet{Zhang2010} show that the skew-$t$ can be expressed as a hierarchical model
\begin{align} \label{eq:fullmodel}
  \bY &= \bX \bbeta + \alpha |z| + \sigma \bv
\end{align}
where $\alpha \in \calR$ controls the skewness; $z \sim N(0, \sigma^2)$ is a random effect from a half-normal distribution (see appendix A.4); $\bv$ is a Gaussian process with mean zero, variance one, \Matern spatial correlation with spatial range $\rho$, smoothness $\nu$, and $\gamma$ is the proportion of variance accounted for by the spatial variation; and $\sigma^2 \sim \text{IG}(a, b)$.
When marginalizing over the $z$ and $\sigma^2$ terms, the joint distribution for all sites becomes
\begin{align*}
  \bY \sim \text{skew-}t (\mu, \bSigma^*, \alpha, \text{df} = 2a)
\end{align*}
where $\mu$ is the location, $\bSigma^* = \frac{ b }{ a } \bSigma$, $\bSigma$ is a \Matern covariance matrix, and $\alpha \in \calR$ controls the skewness.
The skew-$t$ process is desirable because of its flexible tail that is controlled by the skewness parameter, $\alpha$, and the degrees of freedom, $2a$.
Furthermore, if $\bY$ follows a multivariate skew-$t$ distribution, the marginal distributions also follow a skew-$t$ distribution \citep{Azzalini2003}.

\subsection{Extremal dependence}
One measure of extremal dependence is the $\chi$ statistic \citep{Padoan2011}.
The $\chi$ statistic for the upper tail is given by
\begin{align*}
  \chi = \lim_{c \rightarrow \infty} \Pr(Y(\bs_1) > c | Y(\bs_2) > c).
\end{align*}
In a stationary spatial process, we can write the $\chi$ coefficient as
\begin{align*}
  \chi(h) = \lim_{c \rightarrow \infty} \Pr(Y(0) > c | Y(h) > c)
\end{align*}
where $h = ||\bs_1 - \bs_2||$.
If $\chi(h) = 0$, then observations are asymptotically independent at distance $h$.
For Gaussian processes, $\chi(h) = 0$ regardless of the the distance, so they are not suitable for modeling spatially dependent extremes.
However, for the skew-$t$ process described in \ref{s:data}, $\chi(h) > 0$ \citep[see Appendix A.5 for the expression]{Padoan2011}.

\subsection{Random daily partition}\label{s:part}
One problem with the spatial skew-$t$ process is that all sites are asymptotically dependent regardless of their spatial separation.
This occurs because all observations, both near and far, share the same $z$ and $\sigma^2$ terms.
Even more problematic is that $\lim_{h \rightarrow \infty} \chi(h) \neq 0$.
We handle this problem with a random partition similar to \citet{Kim2005} that allows $z$ and $\sigma^2$ to vary by site.

Consider a set of spatial knots $\bw_{k} \sim \text{Uniform}(\calD)$ that define a random daily partition $P_{1}, \ldots, P_{K}$ of the spatial domain of interest $\calD \subset \calR^2$ such that
\begin{align*}
  P_{k} = \{ \bs : k = \argmin_\ell || \bs - \bw_{\ell} || \}.
\end{align*}
Then, the observation at each site becomes
\begin{align}
  Y(\bs) &= X(\bs) \bbeta + \alpha |z(\bs)| + \sigma v(\bs),
\end{align}
and, for $\bs \in P_{k}$,
\begin{align}
  z(\bs) &= z_{tk} \label{eq:sitez}\\
  \sigma^2(\bs) &= \sigma^2_{k} \label{eq:sitesig}.
\end{align}
So, within each partition, $\bY$ follows the distribution given in (\ref{eq:fullmodel}).

When incorporating the random daily partition, the $\chi$ statistic becomes
\begin{align}
  \chi(h) = \pi(h) \chi(h) + (1 - \pi(h)) \chi(h)
\end{align}
where $\pi(h)$ is the probability that two sites separated by distance $h$ are in the same partition.
Asymptotic dependence is eliminated as $h$ increases, because
\begin{align}
  \lim_{h \rightarrow \infty} \chi(h) = \lim_{h \rightarrow \infty} \pi(h) \chi(h) = 0.
\end{align}
A proof of this can be found in Appendix A.3.

Figure \ref{fig:simchi} shows $\chi(h)$ for different sample quantiles from simulated datasets where $\bs \in [0, 1] \times [0, 1]$.
These plots show how the partitioning allows for short range extremal dependence while reducing extremal dependence as the distance between sites increases.
\begin{figure}
  \includegraphics[width=\linewidth]{plots/sim-chi.pdf}
  \caption{$\chi(h)$ for simulated datasets.}
  \label{fig:simchi}
\end{figure}
